Individual and Differential Harm in Redistricting

### 2022-05-02 paper/paper.Rmd
work

```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts. The total voting population for each example is 180. White lines delineate precincts."}
include_graphics(here("paper/figures/minis_schematic.pdf"))
```

@@

```{r harm-mininois, fig.cap="Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b). Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual."}
include_graphics(here("paper/figures/harm_mininois.pdf"))
```

@@

```{r diffharm-minissouri, fig.cap="Three potential districting plans for Minissouris: a Democratic gerrymander (a), a close-to-proportional plan (b), and a Republican gerrymander (c). Panels (d) and (f) show the total harm to voters in each precinct done by their corresponding gerrymander, using (b) as a counterfactual.  Panel (e) summarizes this harm by party and across precincts, showing the average harm done to voters of each party by each gerrymander."}
include_graphics(here("paper/figures/diffharm_minissouri.pdf"))
```

@@

```{r nj-maps, fig.cap="Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan. Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3}
include_graphics(here("paper/figures/nj_maps.pdf"))
```

@@

We sample 10,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

@@

[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].

```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures. Point lightness indicates the number of Democratic seats won at baseline, ranging from 7 (dark) to 10 (light). The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses. The dashed line indicates a value of zero. Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
include_graphics(here("paper/figures/nj_pairs.pdf"))
```

@@

```{r boston-map, fig.cap="Racial composition of the city of Boston (left), and 2010 city coucil districts (right). Neighborhood boundaries are overlaid in black on the left map and in white on the right map.", fig.width=6.75, fig.height=3.375}
include_graphics(here("paper/figures/boston_map.pdf"))
```

@@

```{r boston-factors, fig.cap="Candidate factor loadings, with approximate 80\\% credible intervals shown as ellipses (left), and geographic variation in factor values by race (right). Neighborhood boundaries overlaid in black. Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.", fig.width=7, fig.height=3.5}
include_graphics(here("paper/figures/boston_factors.pdf"))
```

@@

In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^6]
Throughout this appendix, let $j$ index the set of electoral precincts or other units.
For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.

[^6]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
    Population information comes from Census blocks, which are practically never split.

@@

The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^7]
This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.

[^7]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^8]
Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^8]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

@@

# Principal Components Ecological Inference Model for Boston Primary Elections

```{r boston-gen-pred, fig.cap="Model predictions of general election vote shares and turnout, compared to actual general election results. Red reference line indicates perfect agreement.", fig.width=6.5}
include_graphics(here("paper/figures/boston_gen_pred.pdf"))
```

>>>>

```{r mini-schematic}
#| fig.cap: |
#|      Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
#|      The total voting population for each example is 180. 
#|      White lines delineate precincts.
include_graphics("figures/minis_schematic.pdf")
```

@@

```{r harm-mininois}
#| fig.cap: |
#|      Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
#|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
include_graphics("figures/harm_mininois.pdf")
```

@@

```{r diffharm-minissouri}
#| fig.cap: |
#|      Three potential districting plans for Minissouris: a Democratic gerrymander (a), 
#|      a close-to-proportional plan (b), and a Republican gerrymander (c).
#|      Panels (d) and (f) show the total harm to voters in each precinct done 
#|      by their corresponding gerrymander, using (b) as a counterfactual.
#|      Panel (e) summarizes this harm by party and across precincts, showing 
#|      the average harm done to voters of each party by each gerrymander.
include_graphics("figures/diffharm_minissouri.pdf")
```

@@

```{r nj-maps, fig.width=6.5, fig.height=3}
#| fig.cap: |
#|      Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
#|      Census-designated places are shaded in gray and county lines in white.
include_graphics("figures/nj_maps.pdf")
```

@@

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.

@@

```{r nj-pairs, fig.width=7, fig.height=7}
#| fig.cap: |
#|      Scatterplots for all pairs of measures.
#|      The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
#|      The dashed line indicates a value of zero.
#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
include_graphics("figures/nj_pairs.pdf")
```

@@

```{r boston-map, fig.width=6.75, fig.height=3.375}
#| fig.cap: |
#|      Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
#|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
include_graphics("figures/boston_map.pdf")
```

@@

## Approach

To apply our proposed framework to study the council system and and its district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans and systems.
This is more difficult than in our study of New Jersey, for two reasons.
First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
To estimate election outcomes, we will need some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
Second, we aim to examine the racial impact of the district system, and thus need to estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone tell us how different racial groups voted.

```{r boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
#| fig.cap: |
#|      Candidate factor loadings, with approximate 80% credible intervals shown as ellipses (left), 
#|      and geographic variation in factor values by race (right). 
#|      Neighborhood boundaries overlaid in black.
#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
include_graphics("figures/boston_factors.pdf")
```

To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^6] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.

[^6]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.

Like all ecological inference models, ours rests on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not strictly met.
Yet we believe that the model is a useful simplification of reality that captures the major electoral dynamics that are relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.

## Results

@@

In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^7]
Throughout this appendix, let $j$ index the set of electoral precincts or other units.
For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.

[^7]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
    Population information comes from Census blocks, which are practically never split.

@@

The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^8]
This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.

[^8]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.

The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
Specifically, we define $\gamma:[0,1]\to\R$, $$
\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
$$ which is the inverse CDF of a Laplace distribution with rate 2.[^9]
Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.

[^9]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.

@@

# Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}

```{r boston-gen-pred, fig.width=6.5}
#| fig.cap: |
#|      Model predictions of general election vote shares and turnout, compared to actual general election results. 
#|      Red reference line indicates perfect agreement. 
#|      Actual and estimated results have Spearman correlations of 0.87 (vote share) and 0.84 (turnout rate).
include_graphics("figures/boston_gen_pred.pdf")
```
